---
title: "group by and summarize"
output: github_document
---


```{r}
library(tidyverse)


knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp =.6,
  out.width = "90%"
)
```


Loading a dataset

```{r}
weather_df =  
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, everything())
```


`group_by`

```{r}
weather_df %>% 
  group_by(name,month)
```

## count some things

```{r}
weather_df %>% 
  group_by(name,month) %>% 
  summarize(n_obs = n())

```

↑ This is good for doing follow-up analysis, because this gives a pivot-wider format table. Compare with following chunk ↓:

```{r}
weather_df %>% 
  janitor::tabyl(month,name)
```

↑ This gives a table that may be good for results presentation


Yet, don't use this!! (similar to the function of above `tably`)
```{r}
weather_df %>% 
  pull(name) %>% 
  table()
```


```{r}
weather_df %>% 
  janitor::tabyl(month,name)
```

## MOre general summaries

Let's count, take means, look at SDs

```{r}
weather_df %>% 
  # drop_na() %>% 
  group_by(name,month) %>% 
  summarize(
    n_obs = n(),
    mean_tmax = mean(tmax,na.rm = TRUE),
    median_tmax = median(tmax, na.rm = TRUE),
    sd_tman = sd(tmax, na.rm = TRUE)
  ) %>% 
  ggplot(aes(x = month,y = mean_tmax, color = name))+
  geom_point()+
  geom_line()
  
```


Formatting table

```{r}
weather_df %>% 
  group_by(name) %>% 
  summarize(
    n_obs = n(),
    mean_tmax = mean(tmax,na.rm = TRUE)
  ) %>% 
  knitr::kable()

```


## grouped `mutate`

```{r}
weather_df %>% 
  mutate(
    mean_tmax = mean(tmax,na.rm = TRUE)
  )

weather_df %>% 
  group_by(name) %>% 
  mutate(
    mean_tmax = mean(tmax,na.rm = TRUE),
    centered_tmax = tmax - mean_tmax
  ) %>% 
  ggplot(aes(x = date, y = centered_tmax,color = name))+
  geom_point()
  

```

The second add a column mean of `tmax` after group by name. Then it will calculate the mean with each 3 groups, rather than the mean of tmax for the whole data.


```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(
    tmax_rank = min_rank(tmax) ## rank from min to max
  ) %>% 
  filter(tmax_rank<2)

weather_df %>% 
  group_by(name) %>% 
  mutate(
    tmax_rank = min_rank(desc(tmax)) ## rank from max to min
  ) %>% 
  filter(tmax_rank<2)
```


Lagged variables

```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(
    lagged_tmax = lag(tmax,n=1),
    tmax_diff = tmax - lagged_tmax
  ) %>% 
  summarize(
    diff_sd = sd(tmax_diff,na.rm = TRUE)
  )
```


## LImitations



```{r}
weather_df %>% 
  group_by(name) %>% 
  summarize(cor_tmin_tmax = cor(tmin,tmax,use = "complete"))

weather_df %>% 
  filter(name == "CentralPark_NY" )%>% 
           lm(tmax ~ tmin,data = .)
```

However, in this case `group_by()` does not work.

```{r, eval=FALSE}
weather_df %>% 
  group_by(name) %>% 
  summarize(lm = lm(tmax ~ tmin))
```





















